\appendix
\chapter{Installing Gaston}
\begin{intro}

\end{intro}

\section{Installing dependencies}

Before installing the mighty Gaston you need the following friends.

\begin{itemize}
	\item[] git (>= 1.6.0.0)
	\item[] cmake (>= 2.8.2)
	\item[] gcc (>= 4.8.0)
	\item[] flex (>= 2.5.35)
	\item[] bison (>= 2.7.1)
	\item[] python (>= 2.0)
\end{itemize}
\tsf{i'm sure there are actually more of them}

\section{Setup and configure}

In order to compile and run \gaston first clone the source repository:

\begin{lstlisting}[language=bash]
 $ git clone https://github.com/tfiedor/Gaston.git
\end{lstlisting}

Go to source folder and run
\begin{lstlisting}[language=bash]
 $ make release
\end{lstlisting}
to run the Release version of \gaston.

In order to validate the installation and correctness of the tool
run the set of regressive tests:

\begin{lstlisting}[language=bash]
 $ python testbench.py
\end{lstlisting}

\chapter{Syntax of input formulae}

This chapter provides the supported syntax of the Gaston. For full
syntax of MONA formulae syntax conform the official MONA manual~\cite{mona:manual}.

\begin{verbatim}
program ::= (header;)? (declaration;)+
header ::=  ws1s } ws2s

declaration ::= formula
             |  var0 (varname)+
             |  var1 (varname)+
             |  var2 (varname)+
             |  'pred' varname (params)? = formula
             |  'macro' varname (params)? = formula

formula ::= 'true' | 'false' | (formula)
         |  zero-order-var
         | ~formula
         | formula | formula
         | formula & formula
         | formula => formula
         | formula <=> formula
         | first-order-term = first-order-term 
         | first-order-term ~= first-order-term 
         | first-order-term < first-order-term 
         | first-order-term > first-order-term
         | first-order-term <= first-order-term 
         | first-order-term >= first-order-term
         | second-order-term = second-order-term
         | second-order-term = { (int)+ }
         | second-order-term ~= second-order-term
         | second-order-term 'sub' second-order-term
         | first-order-term 'in' second-order-term
         | ex1 (varname)+ : formula
         | all1 (varname)+ : formula
         | ex2 (varname)+ : formula
         | all2 (varname)+ : formula

first-order-term ::= varname | (first-order-term)
                  |  int
                  | first-order-term + int

second-order-term ::= varname | (second-order-term)
                   |  second-order-term + int
\end{verbatim}
\tsf{better representation of the syntax}

\chapter{Command line interface}

\begin{lstlisting}[language=bash]
	Usage: gaston [options] <filename>
\end{lstlisting}

\begin{itemize}
	\item[\texttt{-t}, \texttt{--time}] Prints the elapsed time for each
		phase of the decision procedure
	\item[\texttt{-d}, \texttt{--dump-all}] Prints additional statistics
		about each phases (symbol table, etc.)
	\item[\texttt{-ga}, \texttt{--print-aut}] Outputs the resulting 
		automaton in graphviz format
	\item[\texttt{--no-automaton}] Does not output the resulting
		automaton
	\item[\texttt{--test=OPT}] Tests either satisfiability (\texttt{sat}),
		validity (\texttt{val}) or unsatisfiability (\texttt{unsat}) only
	\item[\texttt{--walk-aut}] Will walk the input formula and try to
		convert each subformula to automaton and prints statistics
	\item[\texttt{-e},\texttt{--expand-tagged}] Expands the automata
		with specific tags that are specified in the first line of
		the input formulae
	\item[\texttt{-q},\texttt{--quite}] Will not print any progress,
		only minimalistic informations
	\item[\texttt{-oX}] Sets optimization level (deprecated)
\end{itemize}

\chapter{Benchmarks}

This appendix sums the benchmarks that are used for evaluation of the tools for WS1S logic.
Every benchmark is briefly described, and quantified with several measures\,---\,number
of occurring variables (both bound and free), number of atomic formulae, and whether it is
valid, satisfiable or unsatisfiable. Note that we do not present some of the other measures
that are used in our experiments, like number of nodes in tree representation of formula, 
number of fixpoint computations and others, as they can vary from the used optimizations
and preprocessing steps.

\section{STRAND: Structure and Data}

This benchmark is obtained from the work of~\cite{strand1, strand2}. These formulae encode
various invariants of the cycles of several chosen algorithms (e.g. insert, delete) over 
various structures (e.g. linked lists). This benchmark is divided into two parts
\strandbenchone and \strandbenchtwo. The first one is the initial attempt to use the
WS1S for deciding the structural invariants, while the latter is the optimization of the 
initial one by reducing the size of the formulae and introducing some of the advanced 
constructs (e.g. predicates) that enabled faster decision procedure.

\begin{table}
 \centering
  \caption{}\label{tab:bench-strand}
  \begin{tabular}{l l r r r}
  \hline
  \textbf{Id} & \textbf{Name} & \textbf{Variables} & \textbf{Atoms} & \textbf{Answer} \\
  \hline
  \hline
  \multicolumn{5}{c}{\strandbenchone}\\
  \hline
  \input{data/bench-strand-old}
  \hline
  \multicolumn{5}{c}{\strandbenchtwo}\\
  \hline
  \input{data/bench-strand-new}
  \hline
  \end{tabular}
\end{table}

\newpage

\section{UABE: Unbounded Arrays Bounded Elements}

This benchmark is provided from the work of the \cite{uabe}. The authors introduced novel theory
of Unbounded Arrays and Bounded Elements allowing one to reason about array structures and its elements.
The proposed logic is shown decidable by reduction to WS1S. The paper provides moderate benchmark
representing invariants of programs manipulating arrays like e.g. sorts, finding minima, etc.

\begin{table}[h!]
 \centering
  \caption{}{\label{tab:bench-uabe}}
  \begin{tabular}{l l r r r}
  \hline
  \textbf{Id} & \textbf{Name} & \textbf{Variables} & \textbf{Atoms} & \textbf{Answer} \\
  \hline
  \hline
  \input{data/bench-uabe}
  \hline
  \end{tabular}
\end{table}

\newpage

\section{Veanes Minimization Benchmark}
This benchmark is selected from the paper~\cite{veanes} that is describing novel minimization
technique, that is said to outperform the technique implemented by MONA. However, as most of the
benchmarks are over symbolic formulae that is not supported by our tool, we only selected those
that are supported by us.

This parametric family of formulae are complex for most of decision procedures, because of
potentially great number of variables (and thus long tracks and number of symbols) and are in
prenex normal form, potentially causing a state explosion. However, we show in Section~\ref{opt:full-ap}
that this can be bypassed by using the technique of anti-prenexing, that greatly reduces the search space.

The family is defined for various $k \geq 2$ as follows:
\begin{equation}
 \exists x_1,\ldots,x_k. \bigwedge_{1 \leq i \leq k - 1} x_i < x_{i+1}
\end{equation}

\begin{table}[h!]
 \centering
  \caption{}{\label{tab:bench-veanes}}
  \begin{tabular}{l l r r r}
  \hline
  \textbf{Id} & \textbf{Name} & \textbf{Variables} & \textbf{Atoms} & \textbf{Answer} \\
  \hline
  \hline
  \input{data/bench-veanes}
  \hline
  \end{tabular}
\end{table}

\section{dWiNA: Set of Generated Formulae}

These benchmarks are provided from the evaluation of our prototype tool \dwina, that implements the
backward decision procedure. These formulae represents several parametric families of formulae
and are in prenex normal form and thus can pose a serious challenge for decision procedures with
the increasing $k$, as the tracks explode.

There are several sets of benchmarks representing some artificial formulae that represents various
properties of set, like e.g. that there exists $k$ sets that are prefix size (\ref{eq:dwina-sets}),
that there exists $k$ singleton sons (\ref{eq:dwina-singletons}) or some very obvious stuff (\ref{eq:obvious}.

\begin{eqnarray}
  \exists X_1,\ldots, X_k. \forall x. \exists y, z. \bigwedge_{1 \leq i \leq k} 
   (x \in X_i \wedge x \leq y \leq z \wedge z \in X_i) \Leftarrow y \in X_i\label{eq:dwina-sets}\\
  \exists X_1,\ldots, X_k. \forall X. \bigwedge_{1 \leq i \leq k} (X \subseteq X_i \wedge X \neq X_i)
    \Leftarrow X_i \not\subseteq X\label{eq:obvious}\\
  \exists X_1,\ldots, X_k. \forall x_1, x_2. \bigwedge_{1 \leq i \leq k}
   (x_1 \in X_i \wedge x_2 \in X_i) \Leftarrow x_1 = x_2\label{eq:dwina-singletons}
\end{eqnarray}

\begin{table}[h!]
 \centering
  \caption{}\label{tab:bench-dwina}
  \begin{tabular}{l l r r r}
  \hline
  \textbf{Id} & \textbf{Name} & \textbf{Variables} & \textbf{Atoms} & \textbf{Answer} \\
  \hline
  \hline
  \multicolumn{5}{c}{\texttt{HORN-TRANS}}\\
  \hline
  \input{data/bench-horn-trans}
  \hline
  \multicolumn{5}{c}{\texttt{SET-CLOSED}}\\
  \hline
  \input{data/bench-set-closed}
  \hline
  \multicolumn{5}{c}{\texttt{SET-OBVIOUS}}\\
  \hline
  \input{data/bench-set-obvious}
  \hline
  \multicolumn{5}{c}{\texttt{SET-SINGLETONS}}\\
  \hline
  \input{data/bench-set-singletons}
  \hline
  \end{tabular}
\end{table}

\newpage

\section{MONA Secrets}

This benchmark comes from the~\cite{mona:secrets} paper, that introduced every significant
optimization that MONA created in order to become efficient.

\begin{table}[h!]
 \centering
  \caption{}{\label{tab:bench-secrets}}
  \begin{tabular}{l l r r r}
  \hline
  \textbf{Id} & \textbf{Name} & \textbf{Variables} & \textbf{Atoms} & \textbf{Answer} \\
  \hline
  \hline
  \input{data/bench-secrets}
  \hline
  \end{tabular}
\end{table}

\section{Regsys}

Regsys consists of several benchmarks, that can be further parametrized by several constants
and model e.g. the optimization problem for production, presburger additions and some
other stuff.

\begin{table}[h!]
 \centering
  \caption{}{\label{tab:bench-regsys}}
  \begin{tabular}{l l r r r}
  \hline
  \textbf{Id} & \textbf{Name} & \textbf{Variables} & \textbf{Atoms} & \textbf{Answer} \\
  \hline
  \hline
  \input{data/bench-regsys}
  \hline
  \end{tabular}
\end{table}
