%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Contents: Who contributed to this Document
% $Id: contrib.tex 533 2015-04-09 13:00:40Z oetiker $
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{small} 
  \noindent Copyright \copyright 2016- Tomas Fiedor and Contributors.  All rights reserved.
 
  This document is free; you can redistribute it and/or modify it
  under the terms of the GNU General Public License as published by
  the Free Software Foundation; either version 2 of the License, or
  (at your option) any later version.
  
  This document is distributed in the hope that it will be useful, but
  \emph{without any warranty}; without even the implied warranty of
  \emph{merchantability} or \emph{fitness for a particular purpose}\@.  See the GNU
  General Public License for more details.
  
  You should have received a copy of the GNU General Public License
  along with this document; if not, write to the Free Software
  Foundation, Inc., 675 Mass Ave, Cambridge, MA 02139, USA.

\end{small}

\chapter{Thank you!}
\noindent The basic skeleton and the template of this manual comes from an english translation of the Austrian \LaTeX book translated by T.Oetiker. We thank the author for having this template open source and we built on this document.

The main contributors of the Gaston tool and its theoretical background are as follows:
\begin{verse}
\contrib{Tomas Fiedor}{ifiedortom@fit.vutbr.cz}{VeriFIT group, BUT FIT}
\contrib{Lukas Holik}{holik@fit.vutbr.cz}{VeriFIT group, BUT FIT}
\contrib{Ondrej Lengal}{lengal@fit.vutbr.cz}{VeriFIT group, BUT FIT}
\contrib{Tomas Vojnar}{vojnar@fit.vutbr.cz}{VeriFIT group, BUT FIT}
\contrib{Petr Janku}{ijanku@fit.vutbr.cz}{VeriFIT group, BUT FIT}
\end{verse}

Moreover we would like to thank to the following people for their valuable insight and remarks that led to the perfection of the
Gaston tool and our published papers.

{\flushleft\small
Dmitriy Traytel,			% <???@??>
Pink~Fluffy~Unicorn,        % <eric@ericabrahamsen.net>
}

\vspace*{\stretch{1}}



\pagebreak
\endinput
%

% Local Variables:
% TeX-master: "lshort2e"
% mode: latex
% mode: flyspell
% End:
